Making Sequential Programs Deterministic

### 2019-04-06 randoop-case-study.tex

We annotated one package at a time, starting with the most central packages
that are most depended upon.  Within a package, we followed a similar
strategy, annotating supertypes first.  If the determinism of classes and methods had been
documented, then our annotation effort
would have been easy, just converting English into type qualifiers.
Unfortunately, this was not the case: we had to reverse-engineer each
specification, largely from the methods it calls.  If a method $m$ calls a
method that we had previously annotated as nondeterministic, we annotated
$m$ as nondeterministic.
Sometimes after we
annotated a subclass, we had to go back and change annotations on
the supertype because the subclass implementation was
inconsistent with the specification we had guessed for the supertype.
After resolving most type-checking warnings by adding or changing
annotations, we manually examined each remaining one to determine whether
it was a bug or a false positive warning.

@@

\begin{itemize}
\item   (22\%)
%      8 constructors guarantee all instances of this class are @Det
  When every instance of a class is \<@Det>, \<@PolyDet> is equivalent to
  \<@Det>, but \theDeterminismChecker does not yet recognize this fact.
\item  (16\%)
%      6 https://github.com/t-rasmud/checker-framework/issues/96
  For a type that cannot be \<@OrderNonDet>, \<@PolyDet("up")> is
  equivalent to \<@PolyDet>.
\item (14\%)
%      5 can't specify receiver for anonymous class
  Java does not have syntax for specifying the type of an anonymous class.
\item (11\%)
%      4 actual type toString returns @PolyDet
  A variable is declared as \<Object> which has nondeterministic \<toString>,
  but at run time the value always has deterministic \<toString>.  For
  example, the run-time value is always an \<Integer> or a \<String>.
\item (5\%)
%      2 iterating over @OrderNonDet collection to modify another
  Iteration over an \<@OrderNonDet> collection in order to make another
  \<@OrderNonDet> collection, but the extracted elements have type
  \<@NonDet>.  For example, the following code is safe, but
  \theDeterminismChecker issues a warning.  Variable \<elt> has type
  \<@NonDet>, so the call to \<add> does not typecheck.
\begin{Verbatim}
  void m(@OrderNonDet List<@Det String> input) {
    @OrderNonDet List<@Det String> output = new List<>();
    for (String elt : input) {  // elt has type @NonDet String
      output.add(elt);          // error: illegal argument
    }
  }
\end{Verbatim}

@@

The first two issues can be resolved by improvements to
\theDeterminismCheckerImplementation.
The third and fourth issues are limitations of the Java language, but by a
local refactoring in the Randoop codebase can make the code typecheck.
Refactoring can also eliminate the last warning.

@@

\todo{Report the number of annotations required, and say a few words about
  the process and the effort required.}

\todo{Easier for someone familiar with the codebase.  Easier on new code.}

It's higher than we would prefer.

@@

One is that we annotated some core components of Randoop to require all
Boolean expressions used in conditionals to be deterministic.  (This is an
option in \theDeterminismCheckerImplementation, which increases soundness;
see \cref{sec:threats}.)  This forced
their clients --- transitively, much of the Randoop codebase --- to be
annotated as deterministic.
These annotations are not incorrect, but they are stricter than necessary
because some method types could otherwise be qualifier-polymorphic.
This would make them usable in more contexts.  It would also reduce the
number of necessary annotations, since qualifier-polymorphism (the
\<@PolyDet> annotation) is the default for
formal parameter and return types.  We plan to re-do the annotation effort
from scratch, and the number of annotations will be much lower.

@@

Even though the number of annotations is higher for our prototype
implementation than we would prefer, it still compares favorably to the
extensive effort by the Randoop developers (\cref{sec:subject-program}),
especially given that that \theDeterminismChecker found issues that the
developers had not.

% LocalWords:  linters preprocess canonicalizing nondeterministing bdeebd
% LocalWords:  e15f9155 b473fd14 deduplication c15ccbf2 ff5b4c4 eda87f src
% LocalWords:  f72234 a8e63fb c975a9f7 e3c56 a4970 a460df97 postprocessing
% LocalWords:  MinimizerTests SpecificationCollection findOverridden Det

>>>>

We annotated one package at a time, starting with the most central packages
that are most depended upon.  Within a package, we followed a similar
strategy, annotating supertypes first.  If the determinism of classes and methods had been
documented, then our annotation effort
would have been easy, just converting English into type qualifiers.
Unfortunately, this was not the case: we had to reverse-engineer each
specification, largely from the methods it calls.  If a method $m$ calls a
method that we had previously annotated as nondeterministic, we annotated
$m$ as nondeterministic.
Sometimes after we
annotated a subclass, we had to go back and change annotations on
the supertype because the subclass implementation was
inconsistent with the specification we had guessed for the supertype.
After resolving most type-checking warnings by adding or changing
annotations, we manually examined each remaining one to determine whether
it was a bug or a false positive warning.
The effort would have been much easier for someone familiar with Randoop,
and yet easier if done while code is being written and is malleable.

@@

\begin{itemize}
\item   (22\%)
%      8 constructors guarantee all instances of this class are @Det
  When every instance of a class is \<@Det>, \<@PolyDet> is equivalent to
  \<@Det>, but \theDeterminismChecker does not yet recognize this fact.
\item  (16\%)
%      6 https://github.com/t-rasmud/checker-framework/issues/96
  For a type that cannot be \<@OrderNonDet>, \<@PolyDet("up")> is
  equivalent to \<@PolyDet>.
\item (14\%)
%      5 can't specify receiver for anonymous class
  Java does not have syntax for specifying the type of an anonymous class.
\item (11\%)
%      4 actual type toString returns @PolyDet
  A variable is declared as \<Object> which has nondeterministic \<toString>,
  but at run time the value always has deterministic \<toString>.  For
  example, the run-time value is always an \<Integer> or a \<String>.
\item (5\%)
%      2 iterating over @OrderNonDet collection to modify another
  Iteration over an \<@OrderNonDet> collection in order to make another
  \<@OrderNonDet> collection, but the extracted elements have type
  \<@NonDet>.  For example, the following code is safe, but
  \theDeterminismChecker issues a warning.  Variable \<elt> has type
  \<@NonDet>, so the call to \<add> does not type check.
\begin{Verbatim}
  void m(@OrderNonDet List<@Det String> input) {
    @OrderNonDet List<@Det String> output = new List<>();
    for (String elt : input) {  // elt has type @NonDet String
      output.add(elt);          // error: illegal argument
    }
  }
\end{Verbatim}

@@

The first two issues can be resolved by improvements to
\theDeterminismCheckerImplementation.
The third and fourth issues are limitations of the Java language, but by a
local refactoring in the Randoop codebase can make the code type check.
Refactoring can also eliminate the last warning.

@@

In Randoop's 24K LOC of code, we wrote
100 \<@NonDet>,
64 \<@OrderNonDet>,
1391 \<@Det>,
and 109 \<@PolyDet> annotations.

The number of \<@Det> annotations --- one per 17 lines of code --- is much
higher than we would prefer.
Nonetheless, it compares favorably to the
extensive effort by the Randoop developers (\cref{sec:subject-program}),
especially given that that \theDeterminismChecker found issues that the
developers had not.

@@

One is that we annotated some core components of Randoop to require all
Boolean expressions used in conditionals to be deterministic.  (This is an
option in \theDeterminismCheckerImplementation, which increases soundness;
see \cref{sec:threats}.)  This forced
their clients --- transitively, much of the Randoop codebase --- to be
annotated as deterministic.
These annotations are not incorrect, but they are stricter than necessary
because some method types could otherwise be qualifier-polymorphic.
This would make them usable in more contexts.  It would also reduce the
number of necessary annotations, since qualifier-polymorphism (the
\<@PolyDet> annotation) is the default for
formal parameter and return types.

@@

In future work, we plan to implement qualifier polymorphism for classes,
then re-do the case study.  We estimate this will reduce the number of
\<@Det> annotations needed by about an order of magnitude.

% LocalWords:  linters preprocess canonicalizing nondeterministing bdeebd
% LocalWords:  e15f9155 b473fd14 deduplication c15ccbf2 ff5b4c4 eda87f src
% LocalWords:  f72234 a8e63fb c975a9f7 e3c56 a4970 a460df97 postprocessing
% LocalWords:  MinimizerTests SpecificationCollection findOverridden Det
% LocalWords:  methodWeights keySet f212cc7e PolyDet OrderNonDet NonDet i7
% LocalWords:  elt LOC
